Migration of Gas Giant Planets in Gravitationally Unstable Disks

### 2011-05-26 Text/migration.tex
Merged in Dick's changes and fixed the last of the issues. Still needs figure tweaking.
	modified:   Text/migration.tex

git-svn-id: http://jaga.astro.indiana.edu/paperrepo/ApJL/Migration2010@80 3b7f7548-5c66-0410-83ab-3997167635fc

Migration of gas giant planets due to interactions with a circumstellar gas disk can play a major role in defining the architecture of planetary systems. Work on migration \citep[see review by][]{papaloizou2007} has included gravitational interaction of planets with both laminar and turbulent disks. However,  the self-gravity of the gas disk has usually been ignored, and the effects of radiative cooling and a detailed equation of state (EOS) have only recently been included \citep[e.g.,][]{paardekooper2006,paardekooper2010,paardekooper2010b}. 
%***DICK: Scott, you need to do this: {\bf NEED TO RECHECK PAARDEKOOPER FOR RADIATIVE COOLING} 
\citet{boss2005} and \citet{mayer2004} examined radial migration of planet-mass fragments in gravitationally unstable disks, but their disks were violently disrupted by fragmentation under conditions (radii $<$ 40 AU, disk-to-star mass ratios $M_d/M_s \sim 0.1$, and stellar mass $M_s \sim 1\;M_{\odot}$) where fragmentation may not actually occur  \citep{rafikov2005,rafikov2007,boley2006,boley2007b,boley2008,forgan2009,cai2010}. More recently, fragmentation into clumps with gas giant or brown dwarf masses has been documented in numerical simulations of disks that are relatively massive ($M_d/M_s \sim$ a few tenths) and spatially extended (outer radii $>$ 50 AU) \citep{krumholz2007,stamatellos2007,stamatellos2009,boley2009,boley2010}, where fragmentation is expected from semi-analytic arguments \citep[e.g.,][]{clarke2009,rafikov2009,dodson2009}. The fate of the clumps then depends in part on their radial migration, which is a chaotic and messy affair in a fragmenting disk \citep[e.g.,][]{boley2009,boley2010,vorobyov2010,boleydurisen2010}. The occurrence of gravitational instabilities (GIs) may also be episodic \citep[e.g.,][]{vorobyov2006,vorobyov2010b,zhu2010}. Clumps that survive and contract to the dimensions of young planets can later find themselves in a disk that erupts again into GI activity. As the star/disk system evolves, such a protoplanet may end up in a region of a GI-active disk where fragmentation does not occur.

@@

{\it The $t = 0$ Planet Run.} Figure \ref{fig:Am} shows that inclusion of even this modest mass planet ($0.7 \%$ of $M_d$) has a dramatic effect on the burst phase. Without a planet, when all modes grow from imposed noise, it takes 4 ORPs for coherent spiral modes to organize and grow to nonlinear amplitude. The growth is centered near the $Q$-minimum at $26$ AU. Modes of cos($m\phi$) symmetry with $m =$ 4 and 5 dominate. An $m = 3$ (three-armed) spiral also grows, but somewhat more slowly, and lags the other modes in time. In the $t = 0$ planet run, the planet develops an organized wake within about an ORP in which $m = 3$ is a dominant component ($\sim 5$ \% global density perturbation). This nonlinear seeding causes $m = 3$ to dominate the GI burst, which also occurs $\sim$ 2 ORP earlier. Because of our initial placement of the planet inside but close to the $Q$-minimum, the corotation radius (CR) of the $m = 3$ mode is fairly close to that of the planet's orbit period. When the triggered $m = 3$ mode becomes strongly nonlinear at about 3 ORP, planet migration is significantly affected. Figure \ref{fig:a} shows the evolution of the planet's radial position. From 0 to 2.5 ORP, the planet is torqued primarily by its own wake and migrates inward. 
%DICK: New words.
Beginning at about 3 ORP, the planet interacts with the now nonlinear $m$ = 3 GI mode. At first, the planet gains angular momentum and moves outward, but, from 4 to 8 ORP, a time interval of  about 720 yr, the plant experiences a negative torque and plunges from 23 to 17 AU.
%***DICK: As we agreed, the wording below could not be verified. I leave it here in case you want to do more analysis to check it out. I have replaced these words with something more vague.
%Beginning at 3 ORP, when the $m = 3$ mode has grown significantly and the planet is lagging one of its arms in azimuth, the planet gains angular momentum and moves outward. At about $t =$ 4 ORP, as the planet overtakes the arm, the sign of the net torque reverses and the planet plunges from 23 to 17 AU in about 4 ORP (10$^3$ yr). During this time, the planet and mode seem to interact strongly enough that the planet essentially surfs along the arm, maintaining a location within the arm as it plunges. 
After $t =$ 8 ORP, the main burst is over, and the disk transitions into its asymptotic state, where modes of many $m$-values become comparably strong (see Fig. \ref{fig:Am}). The planet's radial migration apparently stalls at about 16 to 17 AU. From an analysis of periodicities present in the gas disk between 8 and 12 ORP, the planet lies a few AU inside the inner Lindblad resonance of a strong $m = 2$ mode with CR at 29 AU in this early part of the asymptotic phase.

@@

{\it Eccentricity.} Figure \ref{fig:e} shows the evolution of $e$ in the planet simulations. In both cases, the planets were inserted with an approximate circular velocity computed by adding the interior disk mass to the stellar point mass. The presence of GI activity in the $t = 10$ planet run immediately increases $e$ to $\sim 0.08$. In the $t = 0$ planet run, the modest initial $e$ decreases during the 1.5 ORP when it is migrating only due to its own wake, as expected. 
%***DICK: Scott, we need a reference for this "expectation".
However, once there are strong nonlinear interactions with GI modes, $e$ jumps upward. In both runs, it appears that interaction with well-established GI activity leads to eccentricities between 0.02 and 0.1 that vary in a chaotic way between these extremes on orbit period time scales. The magnitude of $e$ is roughly consistent with the ratio of the sound speed to orbital speed of the disk and the modest values for the Mach numbers of the GI spirals \citep{boley2006a,boley2008}. In other words, the planet's orbit tends to have about as much nonaxisymmetry as reflected in the pitch angles of the GI-induced spirals \citep{cossins2009}.

@@

In both simulations, the inward migration appears to halt near 17 AU. This may simply be a coincidence, because the overall average migration rates are similar, the starting positions are essentially the same, and the calculations have a similar duration. We plan to extend simulations like these to much longer times in the future. However, the final radial positions of both planets happen to be the location of the Inner Lindblad resonance (ILR) for a strong m = 2 mode present in both runs. This ILR is associated with a surface density enhancement in the disk. Near this enhancement, the surface density gradient may be falling steeply enough to halt migration, a mechanism that can work in a laminar disk \citep[note the $\alpha$ dependence in eq. (70) of][]{tanaka2002}. 
%***DICK: Is my reference here OK?.
Alternatively, at 17 AU, the level of GI activity is quite different interior to the planet ($Q \gtrsim 2$) and exterior to the planet ($Q \lesssim 2$), and this may affect the balance of the Lindblad torques exerted by the disk on the planet. Whether the migration really does stall and what the mechanism may be are subjects for further research.

@@

The authors would like to thank F.C. Adams and J.C.B. Papaloizou for useful encouragement in this project. The research was conducted with the support of NASA grants from the Origins of Solar Systems  Program (NNG05GN11G and NNX08AK36G). S.M. was supported during portions of this research by an NESSP Graduate Fellowship and by Indiana Space Grant Consortium Graduate Fellowships; A.C.B. is currently a Sagan Postdoctoral Fellow at U. Florida. This material is based upon work supported by the National Science Foundation under Grants No. ACI-0338618l, OCI-0451237, OCI-0535258, OCI-0504075, and CNS-0521433. Any opinions, findings and conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the views of the National Science Foundation. 
%***DICK: Any other acknowledgments needed? Scott, you had this in here: {\bf need TG acks}

>>>>

Migration of gas giant planets due to interactions with a circumstellar gas disk can play a major role in defining the architecture of planetary systems. Work on migration \citep[see review by][]{papaloizou2007} has included gravitational interaction of planets with both laminar and turbulent disks. However, radiative transport, detailed equations of state (EOS), and the self-gravity of the gas disk have usually been ignored, the effects of a non-isothermal EOS have only recently been included \citep[e.g.,][]{paardekooper2006,paardekooper2010,paardekooper2011}. 
\citet{boss2005} and \citet{mayer2004} examined radial migration of planet-mass fragments in gravitationally unstable disks, but their disks were violently disrupted by fragmentation under conditions (radii $<$ 40 AU, disk-to-star mass ratios $M_d/M_s \sim 0.1$, and stellar mass $M_s \sim 1\;M_{\odot}$) where fragmentation may not actually occur  \citep{rafikov2005,rafikov2007,boley2006,boley2007b,boley2008,forgan2009,cai2010}. More recently, fragmentation into clumps with gas giant or brown dwarf masses has been documented in numerical simulations of disks that are relatively massive ($M_d/M_s \sim$ a few tenths) and spatially extended (outer radii $>$ 50 AU) \citep{krumholz2007,stamatellos2007,stamatellos2009,boley2009,boley2010}, where fragmentation is expected from semi-analytic arguments \citep[e.g.,][]{clarke2009,rafikov2009,dodson2009}. The fate of the clumps then depends in part on their radial migration, which is a chaotic and messy affair in a fragmenting disk \citep[e.g.,][]{boley2009,boley2010,vorobyov2010,boley2010b}. The occurrence of gravitational instabilities (GIs) may also be episodic \citep[e.g.,][]{vorobyov2006,vorobyov2010b,zhu2010}. Clumps that survive and contract to the dimensions of young planets can later find themselves in a disk that erupts again into GI activity. As the star/disk system evolves, such a protoplanet may end up in a region of a GI-active disk where fragmentation does not occur.

@@

{\it The $t = 0$ Planet Run.} Figure \ref{fig:Am} shows that inclusion of even this modest mass planet ($0.7 \%$ of $M_d$) has a dramatic effect on the burst phase. Without a planet, when all modes grow from imposed noise, it takes 4 ORPs for coherent spiral modes to organize and grow to nonlinear amplitude. The growth is centered near the $Q$-minimum at $26$ AU. Modes of cos($m\phi$) symmetry with $m =$ 4 and 5 dominate. An $m = 3$ (three-armed) spiral also grows, but somewhat more slowly, and lags the other modes in time. In the $t = 0$ planet run, the planet develops an organized wake within about an ORP in which $m = 3$ is a dominant component ($\sim 5$ \% global density perturbation). This nonlinear seeding causes $m = 3$ to dominate the GI burst, which also occurs $\sim$ 2 ORP earlier. Because of our initial placement of the planet inside but close to the $Q$-minimum, the corotation radius (CR) of the $m = 3$ mode is fairly close to that of the planet's orbit period. When the triggered $m = 3$ mode becomes strongly nonlinear at about 3 ORP, planet migration is significantly affected. Figure \ref{fig:a} shows the evolution of the planet's radial position. From 0 to 2.5 ORP, the planet is torqued primarily by its own wake and migrates inward. 
Beginning at about 3 ORP, the planet interacts with the now nonlinear $m = 3$ GI mode. At first, the planet gains angular momentum and moves outward, but, from 4 to 8 ORP, a time interval of  about 720 yr, the plant experiences a negative torque and plunges from 23 to 17 AU.

After $t =$ 8 ORP, the main burst is over, and the disk transitions into its asymptotic state, where modes of many $m$-values become comparably strong (see Fig. \ref{fig:Am}). The planet's radial migration apparently stalls at about 16 to 17 AU. From an analysis of periodicities present in the gas disk between 8 and 12 ORP, the planet lies a few AU inside the inner Lindblad resonance of a strong $m = 2$ mode with CR at 29 AU in this early part of the asymptotic phase.

@@

{\it Eccentricity.} Figure \ref{fig:e} shows the evolution of $e$ in the planet simulations. In both cases, the planets were inserted with an approximate circular velocity computed by adding the interior disk mass to the stellar point mass. The presence of GI activity in the $t = 10$ planet run immediately increases $e$ to $\sim 0.08$. In the $t = 0$ planet run, the modest initial $e$ decreases during the 1.5 ORP when it is migrating only due to its own wake, as expected \citep{ward1998,ward2003,goldreich2004}. 
However, once there are strong nonlinear interactions with GI modes, $e$ jumps upward. In both runs, it appears that interaction with well-established GI activity leads to eccentricities between 0.02 and 0.1 that vary in a chaotic way between these extremes on orbit period time scales. The magnitude of $e$ is roughly consistent with the ratio of the sound speed to orbital speed of the disk and the modest values for the Mach numbers of the GI spirals \citep{boley2006a,boley2008}. In other words, the planet's orbit tends to have about as much nonaxisymmetry as reflected in the pitch angles of the GI-induced spirals \citep{cossins2009}.

@@

In both simulations, the inward migration appears to halt near 17 AU. This may simply be a coincidence, because the overall average migration rates are similar, the starting positions are essentially the same, and the calculations have a similar duration. We plan to extend simulations like these to much longer times in the future. However, the final radial positions of both planets happen to be the location of the Inner Lindblad resonance (ILR) for a strong m = 2 mode present in both runs. This ILR is associated with a surface density enhancement in the disk. Near this enhancement, the surface density gradient may be falling steeply enough to halt migration, a mechanism that can work in a laminar disk \citep{paardekooper2009}. 
Alternatively, at 17 AU, the level of GI activity is quite different interior to the planet ($Q \gtrsim 2$) and exterior to the planet ($Q \lesssim 2$), and this may affect the balance of the Lindblad torques exerted by the disk on the planet. Whether the migration really does stall and what the mechanism may be are subjects for further research.

@@

The authors would like to thank F.C. Adams and J.C.B. Papaloizou for useful encouragement in this project. The research was conducted with the support of NASA grants from the Origins of Solar Systems  Program (NNG05GN11G and NNX08AK36G). S.M. was supported during portions of this research by an NESSP Graduate Fellowship and by Indiana Space Grant Consortium Graduate Fellowships; A.C.B. is currently a Sagan Postdoctoral Fellow at U. Florida. This material is based upon work supported by the National Science Foundation under Grants No. ACI-0338618l, OCI-0451237, OCI-0535258, OCI-0504075, and CNS-0521433, and through TeraGrid resources provided by Indiana University, PSC, SDSC, and NCSA. TeraGrid systems are hosted by Indiana University, LONI, NCAR, NCSA, NICS, ORNL, PSC, Purdue University, SDSC, TACC, and UC/ANL.. Any opinions, findings and conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the views of the National Science Foundation.
